Binning Strategies and Related Loss for Large Data

### 2013-04-04 Paper.Rnw
minor edits to figures
Note: I commented out the enormous airline plot that was slowing
compilation and making it too large to send as an attachment and added a
tiny pdf plot as a placeholder

\begin{figure}[hbtp]
  \centering
  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./figure/Binning-DataAndBin.pdf}}
%   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
%   {./figure/RandomBinningDataPoints1}}
  \subfloat[Bin Attribute Diagram]{\includegraphics[keepaspectratio=TRUE, width=4in]
  {./figure/Binning-DataVisualNumericalCenter.pdf}}
  \caption{In illustration (a) points are assigned to bins centered at (5, 25) an (15,25). Points in the standard algorithm are all contained within the bin region. Illustration (b) displays the points assigned to a single bin along with key bin attributes: the visual bin center, ($x_1^\ast$, $y_1^\ast$), and the numerical bin center, ($\bar{x}$,$\bar{y}$). }
  \label{fig:BinningTutorial}
\end{figure}

@@

\begin{figure}[hbtp]
    \centering
		\subfloat[Spatial Loss]{\includegraphics[keepaspectratio=true, width=.54\textwidth]{./figure/Binning-DataVisualNumericalCenter.pdf}}
    \subfloat[Frequency Loss]{\includegraphics[keepaspectratio=true, width=.45\textwidth]{./figure/FreqLoss.pdf}}
\caption{\label{fig:stdBinTutorial2} Types of loss due to binning: (a) The visual center of the bin is different from the numerical center of the bin, leading to two different sources of loss.  (b) Frequency loss associated with a change of bin width to 1 from 0.25 in the example of Figure~\ref{binning}.}   
	\end{figure}

@@

losses <- read.csv("./data/losses.csv", stringsAsFactors=FALSE)
lossmelt <- melt(losses[,c(1:4,6,10)],id=c("Algorithm","x","y"))
lossmelt <- lossmelt[which(lossmelt$y <= 10),]
lossmelt$y <- as.factor(lossmelt$y)
levels(lossmelt$variable) <- c("Numerical Loss","Visual Loss", "Log Frequency Loss")
qplot(x=x, y=value,
      group=interaction(y, Algorithm),
      data=lossmelt, geom="line",
      colour=y, size=I(1),
      xlab="Bin  Width  (Games)", ylab="Proportion Loss")+   
        theme_bw() + theme(legend.position="bottom") + 
        guides(colour = guide_legend(title="Bin Width (Strike Outs)", override.aes = list(size = 2))) +
        facet_grid(variable ~ Algorithm, scales = "free")

@@

\begin{figure}[hbtp]\centering
\subfloat[Million-point sample of Airline Data]{\includegraphics[width=.44\linewidth, keepaspectratio=TRUE]{images/MillionSample.pdf}}\hfil
\subfloat[Minimal binning of Airline Data]{\includegraphics[width=.56\linewidth, keepaspectratio=TRUE]{images/AirlineStdBinning1min.pdf}}
\caption{\label{airline-scatter}Scheduled and actual departure times of  flights across the United States in 2011. The plot on the left is based on a sample of the data, the plot on the right shows all flights. The big patterns are visible in both plots, but the plot on the left misses some of the finer level details in scheduling that is visible in the plot on the right.}
\end{figure}

>>>>

\begin{figure}[hbtp]
  \centering
  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./figure/Binning-DataAndBin.pdf}}
%   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
%   {./figure/RandomBinningDataPoints1}}
  \subfloat[Random Algorithm]{\includegraphics[keepaspectratio=TRUE, width=3in]
  {./figure/RandomBinningDataPoints1.pdf}}
  \caption{Points are assigned to bins centered at (5, 25) an (15,25). Points in the standard algorithm are contained within the defined bin region, where points from the random algorith may be assigned to bins from outside the defined bin regions.}
  \label{fig:BinningTutorial}
\end{figure}

@@

\begin{figure}[hbtp]
    \centering
		\subfloat[Spatial Loss]{\includegraphics[keepaspectratio=true, width=.54\textwidth]{./figure/Binning-DataVisualNumericalCenter.pdf}}
    \subfloat[Frequency Loss]{\includegraphics[keepaspectratio=true, width=.45\textwidth]{./figure/FreqLoss.pdf}}
\caption{\label{fig:stdBinTutorial2} Types of loss due to binning: (a) Numerical loss associated with the cumulative difference from the binned data to the numerical center and visual loss associated with the difference between the numerical and visual bin centers.  (b) Frequency loss associated with a change of bin width to 1 from 0.25 in the example of Figure~\ref{binning}.}   
	\end{figure}

@@

# connect <- dbConnect(dbDriver("MySQL"), user="2009Expo", 
#                      password="R R0cks", port=3306, dbname="baseball", 
#                      host="headnode.stat.iastate.edu")
# pitch <- new("dataDB", co=connect, table="Pitching")
# d1 <- dbData(pitch, vars=c( "G", "SO"))
# bins <- expand.grid(x=c(1:10, 15, 20), y=c(1:5, seq(10, 100, 10)))
# 
# library(multicore)
# lossesRdm <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="random")))
# lossesStd <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="standard")))
# 
# lossesRdm <- cbind(bins, lossesRdm)
# lossesStd <- cbind(bins, lossesStd)
# 
# losses <- rbind(cbind(Algorithm="Random", lossesRdm), cbind(Algorithm="Standard", lossesStd))
# write.csv(losses, "../dbData/data/losses.csv", row.names=FALSE)

losses <- read.csv("./data/losses.csv", stringsAsFactors=FALSE)
lossmelt <- melt(losses[,c(1:4,6,10)],id=c("Algorithm","x","y"))
lossmelt <- lossmelt[which(lossmelt$y == 1 | lossmelt$y == 10 | lossmelt$y == 50),]
lossmelt$y <- as.factor(lossmelt$y)
levels(lossmelt$variable) <- c("Numerical Loss","Visual Loss", "Log Frequency Loss")
qplot(x=x, y=value,
      group=interaction(y, Algorithm),
      data=lossmelt, geom="line",
      colour=y, size=I(1),
      xlab="Bin  Width  (Games)", ylab="Proportion Loss")+   
        theme_bw() + theme(legend.position="bottom") + 
        guides(colour = guide_legend(title="Bin Width (Strike Outs)", override.aes = list(size = 2))) +
        facet_grid(variable ~ Algorithm, scales = "free")

@@

\begin{figure}[hbtp]\centering
%\subfloat[Million-point sample of Airline Data]{\includegraphics[width=.44\linewidth, keepaspectratio=TRUE]{images/MillionSample.pdf}}\hfil
%   Swap Back at later date (shrinks size of pdf to managable)
\subfloat[Million-point sample of Airline Data]{\includegraphics[width=.44\linewidth, keepaspectratio=TRUE]{images/RplotFiller.pdf}}\hfil
\subfloat[Minimal binning of Airline Data]{\includegraphics[width=.56\linewidth, keepaspectratio=TRUE]{images/AirlineStdBinning1min.pdf}}
\caption{\label{airline-scatter}Scheduled and actual departure times of  flights across the United States in 2011. The plot on the left is based on a sample of the data, the plot on the right shows all flights. The big patterns are visible in both plots, but the plot on the left misses some of the finer level details in scheduling that is visible in the plot on the right.}
\end{figure}
